Human capital mediates natural selection in contemporary humans

### 2021-07-16 why-natural-selection.Rmd
New fertility prs; use income_cat not first_job_pay as a control; plan within-sib regressions with/without controls; comments from Abdel; table to explain theory and empirics together.

---
title: "Human capital drives natural selection in contemporary humans"
author: "David Hugh-Jones \\thanks{Corresponding author. School of Economics, University of East Anglia, Norwich, UK. Email: D.Hugh-Jones@uea.ac.uk}, Abdel Abdellaoui \\thanks{Department of Psychiatry, Amsterdam UMC, University of Amsterdam, Amsterdam, The Netherlands. Email: a.abdellaoui@amsterdamumc.nl}"
date: "`r format(Sys.Date(), '%d %B %Y')`"
abstract: "\\noindent Natural selection has been documented in contemporary humans, but  little is known about the mechanisms behind it. We test for natural selection through the association between 33 polygenic scores and fertility, across two generations, using data from UK Biobank (N = 409,629 British subjects with European ancestry). Consistently over time, polygenic scores associated with lower (higher) earnings, education and health are selected for (against). Selection effects are concentrated among lower SES groups, younger parents, people with more lifetime sexual partners, and people not living with a partner. The direction of natural selection is reversed among older parents (22+), or after controlling for age at first live birth. These patterns are in line with economic theories of fertility, in which higher earnings may either increase or decrease fertility via income and substitution effects in the labour market. Studying natural selection can help us understand the genetic architecture of health outcomes: we find evidence in modern day Great Britain for multiple natural selection pressures that vary between subgroups in the direction and strength of their effects, that are strongly related to the socio-economic system, and that may contribute to health inequalities across income groups."
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: false
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
  - \usepackage[format=plain, labelfont={bf,it}, textfont=it]{caption}
  - \usepackage{titlesec}
  - \titleformat*{\section}{\sffamily\LARGE}
  - \titleformat*{\subsection}{\sffamily\itshape\Large}
  - \hypersetup{colorlinks = true, linkcolor = {blue}, linkbordercolor = {white}}
  - \usepackage{amsthm}
  - \theoremstyle{plain}
  - \newtheorem{lem}{\protect\lemmaname}
  - \providecommand{\lemmaname}{Lemma}
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
mainfont: Baskerville
mathfont: Baskerville
sansfont: "Gill Sans"
---

@@

To test directly for natural selection, we regress respondents' number of
children ($N_i$) on their polygenic scores (PGS):

@@

The "selection effect", $\beta$, reflects the strength of natural selection
within the sample. In fact, since polygenic scores are normalized, $\beta$ is
the expected polygenic score among children of the sample
[@beauchamp2016genetic].[^normalization]

@@

To correct for ascertainment bias, we weight participants using population
data. We try three alternative weighting schemes: weighting by geography, age
and presence/absence of a partner; weighting by age and highest educational
qualification; and age, highest qualification, and age at first live birth, for
women only. Figure \@ref(fig:plot-res-wt) plots selection effects among the
entire sample, estimated with the three weighting schemes, and also with uniform
weights. Effects for `r res_sig_all` out of 33 polygenic scores are significant
at $p < 0.05/33$ under all four weighting schemes, with `r res_sig_any` scores
reaching significance under at least one scheme. Weighting makes
a large difference to estimates: mean effect sizes across all
polygenic scores are increased by a factor of `r mean_rel$Geographical`
(geographical weighting), `r mean_rel[["Age/Qualification"]]`
(age/qualification) or `r mean_rel[["Age/Qual/AFLB"]]` (age/qualification/AFLB).
Estimates might be further affected by weighting on other demographic variables.

We also check for purifying and diversifying selection by estimating
\@ref(eq:regression) with a quadratic term. In particular, we find diversifying
selection for educational attainment polygenic scores: at higher values of these
scores, the negative effect on fertility is smaller (Appendix Figure
\@ref(fig:plot-purifying)).

@@

To understand the reasons for this ascertainment bias, and to learn about the
mechanisms underlying natural selection, we split the sample, starting with
basic demographic variables including education, income and sex. These are all
potential sources of ascertainment bias, as mentioned above. We then re-estimate
\@ref(eq:regression) within each subgroup.

We emphasize two points. First, this exercise does
*not* apportion the total correlation with fertility between subgroups. 
That is because the polygenic score may also affect the probability of being in 
each subgroup. For example, the EA3 polygenic score for educational attainment not only affects
fertility among university graduates, and among non-graduates; it also changes
a person's chance of attending university. Second, we do *not* claim that our
observed correlations reflect causal effects of polygenic scores on fertility.
Indeed, polygenic scores correlate with many sources of environmental variation 
that could affect fertility. However, natural selection is a correlational
phenomenon, not a causal one: if a polygenic score correlates with 
low fertility, it is being selected against, whatever the underlying causal
relationship. Irrespective of causality, differences in equation
\@ref(eq:regression) across subgroups can inform us about how natural
selection works, by constraining the set of possible explanations.

@@

\[
U(N)=u(Y)+aN.
\]

Here $a$ captures the strength of preference for children. $u(\cdot)$
is concave and increasing. We treat $N$ as continuous, in line with
the literature: this can be thought of as the expected number of children
among similar people. The marginal benefit of an extra child is
$\frac{dU}{dN} = -bWu'(Y)+a$. The effect of an increase in human capital on this
marginal benefit is

@@

\begin{equation}
U(N_{1},N_{2}) = u(Y_{1}) + u(Y_{2}) + aN_{1} + aN_{2})\label{eq:U}
\end{equation}

Instead of identifying human capital with wages, we now allow
individuals to choose a level of education $s \in [0,1]$, which has
a time cost in period 1. Education is
complementary to human capital $h > 0$, and increases period 2 wages, which take 
the  simple functional form $w(s,h) = sh$. We normalize period 1 wages
to 1. Letting $u(\cdot)$ take the constant relative risk aversion (CRRA) form
$u(y)=\frac{y^{1 - \sigma} - 1}{1 - \sigma}$, where $\sigma>0$ captures risk aversion
in income, we examine total fertility $N^{*} = N_{1}^{*} + N_{2}^{*}$ and the 
*fertility-human capital relationship*, $\frac{dN^{*}}{dh}$. For $\sigma < 1$,
and in a neighbourhood of $\sigma = 1$:

1. The fertility-human capital relationship is negative: $\frac{dN^{*}}{dh} < 0$.
2. The relationship is weaker (closer to zero) at higher wages and/or
  levels of human capital.
3. The relationship is more negative when the time burden of children,
  $b$, is larger.
4. The relationship is weaker at higher levels of education $s$.
5. The relationship is weaker among those who start fertility in period
  2 ($N_{1}^{*} = 0$) than among those who start fertility in period
  1 ($N_{1}^{*} > 0$).

Facts 1-5 parallel our empirical results.[^one-period-results] In this model,
polygenic scores that correlate with human capital would have a negative
correlation with fertility at low income and education levels, and a weaker or
zero correlation at higher levels (Figures \@ref(fig:plot-income-educ-level)a
and \@ref(fig:plot-income-educ-level)b). Among scores which are highly
correlated with human capital, the fertility correlation might also be weaker at
higher levels of the score itself, giving rise to diversifying selection, as we
observe for scores for educational attainment (Appendix Figure
\@ref(fig:plot-purifying)). The correlation would be more negative when the time
burden of children $b$ is high, e.g. for single parents (Figure
\@ref(fig:plot-n-partners)). Lastly, the correlation would be weaker among those
starting fertility later (Figure \@ref(fig:plot-age-flb)).[^money-costs]

[^one-period-results]: Facts 1-3 also hold in the one-period model.

[^money-costs]: Our results are actually stronger than this, in that
correlations with fertility are reversed at higher AFLB. This prediction can be
accommodated in the model if children have a money cost as well as a time cost. 
Appendix Figure \@ref(fig:N-plot-with-m) shows an example.

@@

The above does not aim to provide a complete theory of fertility. Rather, it
shows that a relatively simple economic model can explain many of our results.
Other empirical work in economics also supports our mechanism.
@caucutt2002women and @monstad2008education show that education and skills
affect age at first birth and fertility. Income decreases fertility at low
income levels, but increases it at higher income levels [@cohen2013financial].
US fertility decreases faster with education among single mothers than married
mothers [@baudin2015fertility], in line with our fact 3 and as predicted by
@becker1981treatise.

@@

Second, we re-estimate equation \@ref(eq:regression) for each polygenic score,
controlling for earnings in first job and education levels. Effect sizes are
generally reduced (Appendix Figure \@ref(fig:plot-res-earn-educ)), and only 
`r n_05_33` out of 33 scores are significant at $p < 0.05/33$. This suggests that
earnings and education indeed mediate the effect of polygenic scores on
fertility.

@@

An alternative theory is that polygenic scores correlate with the motivation to
have children (parameter $a$ in the model; cf. @jones2008fertility). However, this
theory would not explain why our results are weaker at higher incomes and
education levels.[^note] Another alternative is that traits under selection are
linked to externalizing behaviour and risk-seeking. This might be partially
captured by our risk-aversion parameter $\sigma$, which affects fertility by
changing the marginal utility of income, but a more direct channel is risky
sexual behaviour [@Mills2020.05.06.081273]. The data here provide some support
for this: scores which might plausibly be linked to externalizing behaviour,
like ADHD and (younger) age at smoking initiation, are selected for. However,
this theory is less good at explaining variation in selection across the full
range of scores, including physical measures such as waist-hip ratio and BMI. We
test this theory directly by re-estimating equation \@ref(eq:regression)
controlling for a measure of risk attitude (UK Biobank field 2040).  The median
ratio of effect sizes between regressions with and without controls is 
`r median(ratio)`; all scores which are significant at $p < 0.05/33$ in
uncontrolled regressions remain so when controlling for risk attitude. This
non-result could simply reflect the imprecision of the risk aversion measure,
which is a single yes/no question. However, the risk aversion measure *does*
predict the overall number of children, highly significantly ($p < 2 \times
10^{-16}$, `r n_risk_sig`/33 regressions). Given that, and the statistical power
we get from our sample size, we believe that the non-result is real: while risk
aversion does predict fertility in the sample, it is not an important channel
for natural selection.

@@

Polygenic scores which correlate with high (low) earnings and more (less)
education are being selected against (selected for). In addition, many of the
phenotypes under positive selection are linked to disease risk. Most people
would probably prefer to have high educational attainment, a low risk of ADHD
and major depressive disorder, and a low risk of coronary artery disease, but
natural selection is pushing against genes associated with these traits.
Potentially, this could increase the health burden on modern populations, but
that depends on effect sizes. Our results suggest that naïve estimates can be
affected by sample ascertainment bias. This problem may be less serious in
surveys which aim to be representative (as the UK Biobank does not). However,
there is still scope for bias, since not all respondents consent to the
collection of genetic data. For instance, completion rates for genotype data in
the US Health and Retirement Study were around 80-85% [@hrs2020]. Researchers
should be aware of the risks of ascertainment when studying modern natural
selection.

@@

Because selection effects are concentrated in lower-income groups, they may also
increase inequality with respect to polygenic scores. For example, Figure
\@ref(fig:plot-mean-EA3-by-income) graphs mean polygenic scores for educational
attainment (EA3) among children from households of different income groups. The
blue bars show the actual means, i.e. parents' mean polygenic score weighted by
number of children. The grey bars show the hypothetical means if all households
had equal numbers of children. Natural selection against genes associated with
educational attainment, which lowers the mean, is stronger at the bottom of the
income distribution, and this increases the differences between groups. Overall,
natural selection increases inequality for `r n_increases` out of 33 polygenic
scores, with a median increase of `r median_pct_change`% in the difference
between highest and lowest income groups (Appendix Table
\@ref(tab:tbl-inequality-effects)). Inequalities in polygenic scores are
believed to be important for understanding class mobility and structure 
[see e.g. @belsky2018genetic, @Rimfeld_2018]. If so, these changes are substantial
effects. Since many polygenic scores are predictive of disease risk, they could
also potentially increase health inequalities. In general, the evolutionary
history of anatomically modern humans is related to disease risk
[@benton2021influence]; understanding the role of contemporary natural selection
may aid in mapping the genetic architecture of current health disparities.

@@

\FloatBarrier
\clearpage
## Purifying and diversifying selection

We rerun equation \@ref(eq:regression), adding a quadratic term in $PGS_i$,
and using our age/qualification weights. Scores for hip circumference and drinks
per week show significant purifying selection ($p < 0.05/33$, negative
coefficient). Scores for educational attainment (EA2 and EA3) show significant
diversifying selection ($p < 0.05/33$, positive coefficient), which reduces the
strength of selection against educational attainment at very high levels of the 
PGS. Figure
\@ref(fig:plot-purifying) plots predicted number of children against polygenic
score from these regressions.

We also checked for purifying selection in the parents' generation, using
age/qualification weights multipled by number of siblings. Scores for EA2 and
EA3 again show significant diversifying selection ($p < 0.05/33$, positive coefficient). Other scores including hip circumference and drinks per week 
are not significant.

>>>>

---
title: "Human capital mediates natural selection in contemporary humans"
author: "David Hugh-Jones \\thanks{Corresponding author. School of Economics, University of East Anglia, Norwich, UK. Email: D.Hugh-Jones@uea.ac.uk}, Abdel Abdellaoui \\thanks{Department of Psychiatry, Amsterdam UMC, University of Amsterdam, Amsterdam, The Netherlands.}"
date: "`r format(Sys.Date(), '%d %B %Y')`"
abstract: "\\noindent Natural selection has been documented in contemporary humans, but  little is known about the mechanisms behind it. We test for natural selection through the association between 33 polygenic scores and fertility, across two generations, using data from UK Biobank (N = 409,629 British subjects with European ancestry). Consistently over time, polygenic scores associated with lower (higher) earnings, education and health are selected for (against). Selection effects are concentrated among lower SES groups, younger parents, people with more lifetime sexual partners, and people not living with a partner. The direction of natural selection is reversed among older parents (22+), or after controlling for age at first live birth. These patterns are in line with economic theories of fertility, in which higher earnings may either increase or decrease fertility via income and substitution effects in the labour market. Studying natural selection can help us understand the genetic architecture of health outcomes: we find evidence in modern day Great Britain for multiple natural selection pressures that vary between subgroups in the direction and strength of their effects, that are strongly related to the socio-economic system, and that may contribute to health inequalities across income groups."
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: false
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
  - \usepackage[format=plain, labelfont={bf,it}, textfont=it]{caption}
  - \usepackage{titlesec}
  - \titleformat*{\section}{\sffamily\LARGE}
  - \titleformat*{\subsection}{\sffamily\itshape\Large}
  - \hypersetup{colorlinks = true, linkcolor = {blue}, linkbordercolor = {white}}
  - \usepackage{amsthm}
  - \theoremstyle{plain}
  - \newtheorem{lem}{\protect\lemmaname}
  - \providecommand{\lemmaname}{Lemma}
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
mainfont: Baskerville
mathfont: Baskerville
sansfont: "Gill Sans"
---

@@

To test more directly for natural selection, we regress respondents' number of
children ($N_i$) on their polygenic scores (PGS):

@@

The "selection effect", $\beta$, reflects the strength of natural selection
within the sample. In fact, since polygenic scores are normalized, $\beta$ is
the expected polygenic score among children of the sample
[@beauchamp2016genetic].[^normalization] Note that equation 
\eqref{eq:regression} does not include many possible environmental and
genetic confounds that could affect fertility, and as a result, $\beta$ is not
an estimate of the causal effect of a polygenic score on fertility. However, 
natural selection is a correlational phenomenon, not a causal one: if a 
polygenic score correlates with low fertility, it is being selected against, 
whatever the underlying causal relationship.

@@

To correct for ascertainment bias, we weight participants using population
data. We try three alternative weighting schemes: (1) "Geographical" weighting,
by geography, age and presence/absence of a partner; (2) age and highest educational
qualification; and (3) for women only, age, highest qualification, and age at 
first live birth. Figure \@ref(fig:plot-res-wt) plots selection effects among the
entire sample, estimated with the three weighting schemes, and also with uniform
weights. Effects for `r res_sig_all` out of 33 polygenic scores are significant
at $p < 0.05/33$ under all four weighting schemes, with `r res_sig_any` scores
reaching significance under at least one scheme. Weighting makes
a large difference to estimates: mean effect sizes across all
polygenic scores are increased by a factor of `r mean_rel$Geographical`
(geographical weighting), `r mean_rel[["Age/Qualification"]]`
(age/qualification) or `r mean_rel[["Age/Qual/AFLB"]]` (age/qualification/AFLB).
Estimates might be further affected by weighting on other demographic variables.

We also check for balancing and diversifying selection by estimating
\@ref(eq:regression) with a quadratic term. In particular, we find diversifying
selection for educational attainment polygenic scores: at higher values of these
scores, the negative effect on fertility is smaller (Appendix Figure
\@ref(fig:plot-purifying)).

@@

To understand the reasons for the ascertainment bias in our sample, and to learn 
about the mechanisms underlying natural selection, we split the sample, starting with
basic demographic variables including education, income and sex. These are all
potential sources of ascertainment bias, as mentioned above. We then re-estimate
\@ref(eq:regression) within each subgroup.

We emphasize that this exercise does
*not* apportion the total correlation with fertility into subgroups. 
That is because the polygenic score may also affect the probability of being in 
each subgroup. For example, the EA3 polygenic score for educational attainment not only affects
fertility among university graduates, and among non-graduates; it also changes
a person's chance of attending university. Nevertheless, differences in equation
\@ref(eq:regression) across subgroups can inform us about how natural
selection works, by constraining the set of possible mechanisms.

@@

\[
U = u(Y)+aN.
\]

Here $a$ captures the strength of preference for children. $u(\cdot)$ captures
the taste for income, and is concave and increasing. We treat $N$ as continuous, in line with
the literature: this can be thought of as the expected number of children
among similar people. The marginal benefit of an extra child is
$\frac{dU}{dN} = -bWu'(Y)+a$. The effect of an increase in human capital on this
marginal benefit is

@@

\begin{equation}
U(N_{1},N_{2}) = u(Y_{1}) + u(Y_{2}) + aN_{1} + aN_{2}\label{eq:U}
\end{equation}

Instead of identifying human capital with wages, we now allow
individuals to choose a level of education $s \in [0,1]$, which has
a time cost in period 1. Education is
complementary to human capital $h > 0$, and increases period 2 wages, which take 
the  simple functional form $w(s,h) = sh$. We normalize period 1 wages
to 1. Letting $u(\cdot)$ take the constant relative risk aversion (CRRA) form
$u(y)=\frac{y^{1 - \sigma} - 1}{1 - \sigma}$, where $\sigma>0$ captures risk aversion
in income, we examine total fertility $N^{*} = N_{1}^{*} + N_{2}^{*}$ and the 
*fertility-human capital relationship*, $\frac{dN^{*}}{dh}$. For $\sigma < 1$,
and in a neighbourhood of $\sigma = 1$, Table \@ref(tab:theory) shows five theoretical 
predictions, along with corresponding empirical results for the correlation between
polygenic scores and fertility.[^one-period-results]

[^one-period-results]: Predictions 1-3 also hold in the one-period model. Our empirical 
results are actually stronger than prediction 5, in that
correlations with fertility are *reversed* at higher AFLB. This prediction can be
accommodated in the model if children have a money cost as well as a time cost. 
Appendix Figure \@ref(fig:N-plot-with-m) shows an example.

------------------------------------------------------------------------------------
    **Theory**                                        **Empirical results**
--- ------------------------------------------------- ------------------------------
1.  The fertility-human capital                       Figures \@ref(fig:plot-means-over-time) 
    relationship is negative:                         and \@ref(fig:plot-res-wt).
    $\frac{dN^{*}}{dh} < 0$.

2.  The fertility-human capital                       Figure \@ref(fig:plot-income-educ-level)a. 
    relationship is weaker (closer to zero)           Selection effects are also weaker at higher
    at higher wages and/or levels of                  polygenic scores for EA (Appendix Figure
    human capital.                                    \@ref(fig:plot-purifying)).

3.  The relationship is more negative when            Stronger effects for single parents
    the time burden of children, $b$, is              (Figure \@ref(fig:plot-n-partners)).
    larger.

4.  The relationship is weaker at higher              Figure \@ref(fig:plot-income-educ-level)b.
    levels of education $s$.

5.  The relationship is weaker among those            Effects weaker among those starting 
    who start fertility in period 2                   fertility later 
    ($N_{1}^{*} = 0$) than among those who            (Figure \@ref(fig:plot-age-flb)a).
    start fertility in period 1 
    ($N_{1}^{*} > 0$).
----------------------------------------------------------------------------------------

Table: (\#tab:theory) Predictions from the theoretical model and corresponding 
empirical results.

@@

The above does not aim to provide a complete theory of fertility. Rather, it
shows that a relatively simple economic model can explain many of our results.
Other empirical work in economics also supports our mechanism.
@caucutt2002women and @monstad2008education show that education and skills
affect age at first birth and fertility. Income decreases fertility at low
income levels, but increases it at higher income levels [@cohen2013financial].
US fertility decreases faster with education among single mothers than married
mothers [@baudin2015fertility], in line with our prediction 3 and as predicted by
@becker1981treatise.

@@

Second, we re-estimate equation \@ref(eq:regression) for each polygenic score,
controlling for earnings in first job and education levels. Effect sizes are
generally reduced (Appendix Figure \@ref(fig:plot-res-earn-educ)), with
large reductions for polygenic scores for educational attainment. This 
suggests that earnings and education indeed mediate the effect of 
polygenic scores on fertility.

* We estimate the causal effect of each PGS on fertility using within-sibling
  regressions.
  - These are causal due to lottery of meiosis
  - Also do the fertility PGS

* We also estimate the effect of PSEA and fertility PGS on education and
  earnings, again within-siblings
  - confirm PSEA affects both; does fertility? What if we put both in together?
* Lastly we again estimate effect of PSEAs on fertility, 
  controlling for education and earnings in first job. 
  - We cannot exclude some reverse causality from EA3 -> fertility -> education/income
  - Our theory indeed assumes that people simultaneously choose N and s, while taking into
    account their effect on Y. We don't make claims about exogenous (chance) variation
    in Y or in N. But note that exog variation in N1 (i.e. no choice, treat
    it as fixed to different values) would affect s. OTOH, if EA3 affects $a$ and
    thereby changes things, you don't get predictions 2 or 4.

* Results:
  - in bivariate within-sib regs, extraversion, EA2/3, ADHD and some others are
    weakly significant (only at .05!)
  - adding education as a control barely changes effect sizes
  - adding income as a control increases effect sizes (of EA at least)
  - by contrast, in pooled regs, adding education shrinks effect sizes, and
    adding income does nothing.
  - relevant: effect of EA2 on education is halved within-sib, and effect of
    education on fertility is halved within-sib
  - whereas, effect of EA2 on income is halved within-sib, but effect of income
    on fertility is multiplied by 10!! within-sib. Much of this is because
    of year of birth confound i.e. declining fertility and also retired people
    have smaller income

@@

An alternative theory is that polygenic scores correlate with the motivation to
have children (parameter $a$ in the model; cf. @jones2008fertility). However, this
theory would not explain why our results are weaker at higher incomes and
education levels.[^note] Another alternative is that traits under selection are
linked to externalizing behaviour and risk-seeking. This might be partially
captured by our risk-aversion parameter $\sigma$, which affects fertility by
changing the marginal utility of income, but a more direct channel is risky
sexual behaviour [@mills2021identification]. The data here provide some support
for this: scores which might plausibly be linked to externalizing behaviour,
like ADHD and (younger) age at smoking initiation, are selected for. However,
this theory is less good at explaining variation in selection across the full
range of scores, including physical measures such as waist-hip ratio and BMI. We
test this theory directly by re-estimating equation \@ref(eq:regression)
controlling for a measure of risk attitude (UK Biobank field 2040).  The median
ratio of effect sizes between regressions with and without controls is 
`r median(ratio)`; all scores which are significant at $p < 0.05/33$ in
uncontrolled regressions remain so when controlling for risk attitude. This
non-result could simply reflect the imprecision of the risk aversion measure,
which is a single yes/no question. However, the risk aversion measure *does*
predict the overall number of children, highly significantly ($p < 2 \times
10^{-16}$, `r n_risk_sig`/33 regressions). Given that, and the statistical power
we get from our sample size, we believe that the non-result is real: while risk
aversion does predict fertility in the sample, it is not an important channel
for natural selection.

@@

Polygenic scores which correlate with high (low) earnings and more (less)
education are being selected against (selected for). In addition, many of the
phenotypes under positive selection are linked to disease risk. Many people
would probably prefer to have high educational attainment, a low risk of ADHD
and major depressive disorder, and a low risk of coronary artery disease, but
natural selection is pushing against genes associated with these traits.
Potentially, this could increase the health burden on modern populations, but
that depends on effect sizes. Our results suggest that naïve estimates can be
affected by sample ascertainment bias. This problem may be less serious in
surveys which aim to be representative (as the UK Biobank does not). However,
there is still scope for bias, since not all respondents consent to the
collection of genetic data. For instance, completion rates for genotype data in
the US Health and Retirement Study were around 80-85% [@hrs2020]. Researchers
should be aware of the risks of ascertainment when studying modern natural
selection.

@@

Because selection effects are concentrated in lower-income groups, they may also
increase inequality with respect to polygenic scores. For example, Figure
\@ref(fig:plot-mean-EA3-by-income) graphs mean polygenic scores for educational
attainment (EA3) among children from households of different income groups. The
blue bars show the actual means, i.e. parents' mean polygenic score weighted by
number of children. The grey bars show the hypothetical means if all households
had equal numbers of children. Natural selection against genes associated with
educational attainment, which lowers the mean, is stronger at the bottom of the
income distribution, and this increases the differences between groups. Overall,
natural selection increases inequality for `r n_increases` out of 33 polygenic
scores, with a median increase of `r median_pct_change`% in the difference
between highest and lowest income groups (Appendix Table
\@ref(tab:tbl-inequality-effects)). If inequalities in polygenic scores are
important for understanding class mobility and structure [@belsky2018genetic;
@Rimfeld_2018], then these increases are substantial effects. Since many
polygenic scores are predictive of disease risk, they could also potentially
increase health inequalities. In general, the evolutionary history of
anatomically modern humans is related to disease risk [@benton2021influence];
understanding the role of contemporary natural selection may aid in mapping the
genetic architecture of current health disparities.

@@

\FloatBarrier
\clearpage
## Balancing and diversifying selection

We rerun equation \@ref(eq:regression), adding a quadratic term in $PGS_i$,
and using our age/qualification weights. Scores for hip circumference and drinks
per week show significant balancing selection ($p < 0.05/33$, negative
coefficient). Scores for educational attainment (EA2 and EA3) show significant
diversifying selection ($p < 0.05/33$, positive coefficient), which reduces the
strength of selection against educational attainment at very high levels of the 
PGS. Figure
\@ref(fig:plot-purifying) plots predicted number of children against polygenic
score from these regressions.

We also checked for balancing selection in the parents' generation, using
age/qualification weights multiplied by the inverse of number of siblings. 
Scores for EA2 and EA3 again show significant diversifying selection 
($p < 0.05/33$, positive coefficient). Other scores including hip circumference 
and drinks per week are not significant.
